---
output: html_document
---

#Title: "Programming Assignment 1 Reproducible Research Course"

###Author: "HGTHAKOR"
###Date: "May 28, 2016"
output: html_document
===========================================

###Introduction:

This document is prepared for submission of the Programming Assignment for the "Reproducible Reseach Course". The data provided in the zip file "Activity" from the course is used to analyze and answer the questions given in the course assignment. 

The data from personal activity monitoring device are given. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an an individual collected during the months of October and November, 2012. It includes the number of steps taken in 5 minute intervals on each day.

###Preparation of R

In this document, the analysis and results are written in a single R markdown document that can be processed by knitr and transformed into an HTML file.

It also shows the codes which have been used to derive the results and it has been Set the default of echo to be true throughout the document as follow:

```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
```

###Q. 1 Loading of packages:

###Ans.1:

Following packages have been  loaded for analysis:
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

###Loading and preprocessing the data:

The data is loaded in the working directory with following code and then processed using differnt codes to answer the given questions:

```{r}
getwd()
```

###Data Reading:

###The data is loaded using the read.csv code

```{r}
data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character", "integer"))
```

```{r}
str(data)
```

```{r}
head(data)
```

###Tyding the data
The date is changed into dateformat using lubridate:

```{r}
data$date <- ymd(data$date)
```

Checking the data with str(data)  head(data)

```{r}
str(data)
```

```{r}
head(data)
```

###After tyding the data, it is ready for analysis and the answers to the questions are calculated

### Q.2. What is mean total number of steps taken per day?

### Ans.2: 
There are some missing values, which have been ignored and the steps are:
1. First, calculate the total number of steps taken per day.
2. Second, make a histogram of the total number of steps taken each day.
3. Third, Calculate and report the mean and median of the total number of steps taken per day.

1. Calculate total number of steps taken per day using deplyr and group by date:

```{r}
steps <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
``` 

2. Use ggplot for making the hisogram

```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "red", binwidth = 1000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
```


3.Calculate and report the mean and median of the total number of steps taken per day.

```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)
mean_steps
```

```{r}
median_steps <- median(steps$steps, na.rm = TRUE)
median_steps
```

So, Mean steps are 10766 and median steps are 10765


     
###Q.3: What is the average daily activity pattern?

a.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

###Ans. 3:
First Calculate the average number of steps taken in each 5-minute interval per day using dplyr and group by interval.

```{r}
interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
```

Then Use ggplot for making the time series of the 5-minute interval and average steps taken as follows:

```{r} 
ggplot(interval, aes(x=interval, y=steps)) +   geom_line(color = "blue")
```

Use "which.max()" to find out the maximum steps, on average, across all the days

```{r}
interval[which.max(interval$steps),]
```

It shows that interval 835 is having maximum steps on average amongst all intervals with counts of 206.

###Q.4 Imputing missing values:

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. Devise a strategy for filling in all of the missing values in the dataset. 
3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

###Ans. 4:
1. Summarizing all missing values:

```{r}
sum(is.na(data$steps))
```

So. there are 2304 missing values

2.Taking the approach to fill in a missing NA with the average number of steps in the same 5-min interval

3.Creating  a new dataset as the original and use tapply for filling in the missing values with the average number of steps per 5-minute interval:

```{r}
data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

Checking that there are no missing values in data_full

```{r}
sum(is.na(data_full$steps))
```

So, now there are no missing values.

4. Calculating the number of steps taken in each 5-minute interval per day using "dplyr"" and group by "interval". Using ggplot for making the histogram as follows:

```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

```{r}
ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "red", binwidth = 1000) +
  labs(title = "Histogram of Steps per day, including missing values", x = "Steps per  day", y = "Frequency")
```  

Calculating the mean and median steps with the filled in values:  

```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
mean_steps_full
```

```{r}
median_steps_full <- median(steps_full$steps, na.rm = TRUE)
median_steps_full
```

The results shows no difference between mean and median in new data. The impact of imputing missing data with the average number of steps in the same 5-min interval is that both the mean and the median are equal to the same value: 10766.


###Q.5 Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

###Ans. 5:
1. First use "dplyr"" and "mutate"" to create a new column, weektype, and apply whether the day is weekend or weekday:

```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)
```

Calculate the average steps in the 5-minute interval and use ggplot for making the time series of the 5-minute interval for weekday and weekend, and compare the average steps:

```{r}
interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
```

```{r}
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```

The graphs show that the test object is more active earlier in the day during weekdays compared to weekends, but more active throughout the weekends compared with weekdays, may be because the oject is working during the weekdays, hence moving less during the day time).








